---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 6 @ 11:59PM
author: Tomoki Okuno
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0"
}
```

In this exercise, we use tidyverse (ggpot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the tidyverse.

**Answer:** I used `system.time` to measure run times for the three functions.
```{r}
system.time(tmp_r <- read.csv("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
system.time(tmp_t <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
system.time(tmp_f <- fread("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
```
Looking at user time, CPU time charged for the execution of user instructions, `fread` is the fastest function. XXX

I also used `str` to check data types for them.
```{r}
str(tmp_r)
str(tmp_t)
str(tmp_f)
```


## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c("/mnt/mimiciv/1.0/", "/icu/icustays.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`.  
**Answer:** I used `read_csv` to import this file as a tibble.
```{r}
icustays_tble <- read_csv("/mnt/mimiciv/1.0/icu/icustays.csv.gz")
```

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays?  
**Answer:** There are 53,150 unique `subject_id`s, which is less than the total number of rows. Hence, `subject_id` can have multiple ICU stays.
```{r}
subject_ids <- icustays_tble %>%
  group_by(subject_id) %>%
  summarise(n())
subject_ids
```

3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.  
**Answer:** I used `arrange` to sort `intimes` in ascending order, and then I removed the duplicates for `subject_id`. As a result, the new table has 53,150 rows.  
*Question: we also have to keep the name (or overwrite)?*
```{r}
icustays_tble <- icustays_tble %>%
  arrange(subject_id, intime)  %>%
  distinct(subject_id, .keep_all = TRUE)
icustays_tble
```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.  
**Answer:** I used `read_csv` to import this file as a tibble.
```{r}
admissions_tble <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
```

2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.
```{r}
admissions_tble <- admissions_tble %>%
  semi_join(icustays_tble, by=c("subject_id", "hadm_id"))
```

3. Summarize the following variables by graphics. 

- admission year  
- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  
```{r}
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = year(admittime)))
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = lubridate::month(admittime, label=T)))
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = mday(admittime)))
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = lubridate::wday(admittime, label=T)))
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = hour(admittime)))
```


## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).  
**Answer:** I used `read_csv` to import this file as a tibble.
```{r}
patients_tble <- read_csv("/mnt/mimiciv/1.0/core/patients.csv.gz")
patients_tble <- patients_tble %>% 
  semi_join(icustays_tble, by="subject_id")
```

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents_filtered_itemid.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.  
**Answer:** `labevents.csv.gz` has 122,103,667 rows. (`labevents_filtered_itemid.csv.gz` has 30,198,826 rows.)
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | tail -n +2 | wc -l"
  ), 
  intern = TRUE
)
```

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`.  

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.
    
**Answer:** I used `read_`csv` to import this file as a tibble.
```{r}
labevents_tble <- read_csv("/mnt/mimiciv/1.0/hosp/labevents_filtered_itemid.csv.gz", lazy = TRUE)

# Now we do not have to use col_select? The original file has 15 columns.
choice <- c("50912", "50971", "50983", "50902", "50882", "51221", "51301", "50931", "50960", "50893")

labevents_tble <- labevents_tble %>%
  filter(itemid %in% choice) %>%
  semi_join(icustays_tble, by="subject_id")
labevents_tble
```

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 
```{r}
icustays_tble
labevents_tble <- labevents_tble %>%
  arrange(subject_id, charttime)  %>%
  distinct(subject_id, .keep_all = TRUE) #53097
labevents_tble
```

4. Summarize the lab measurements by appropriate numerics and graphics.
```{r}
labevents_tble
ggplot(data = labevents_tble) + 
  geom_bar(mapping = aes(x = itemid), na.rm = TRUE)
```


## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit

