---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 6 @ 11:59PM
author: Tomoki Okuno
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/tomokiokuno/mimic-iv-1.0"
}
```

In this exercise, we use tidyverse (ggpot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the tidyverse.

**Solution:** I used `system.time` to measure run times for the three functions.
```{r}
system.time(tmp_r <- read.csv(str_c(mimic_path,"/core/admissions.csv.gz")))
system.time(tmp_t <- read_csv(str_c(mimic_path,"/core/admissions.csv.gz")))
system.time(tmp_f <- fread(str_c(mimic_path,"/core/admissions.csv.gz")))
```
Looking at user time, CPU time charged for the execution of user instructions, `fread` is the fastest function.

Next, I used `str` to confirm data types for them. Apparently, there are differences in the (default) parsed data types. For instance, `read.csv` and `fread` read `subject_id` as `integer` while `read_csv` read this as  (double). Also, `read_csv` and `fread` imported the date and time as `POSIXct`, but `read.csv` imported as `factor`.
```{r}
str(tmp_r)
str(tmp_t)
str(tmp_f)
```


## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/icustays.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`.  
**Solution:** I used `read_csv` to import this file as a tibble. The total number of rows is 76,540.
```{r}
icustays_master <- read_csv(str_c(mimic_path, "/icu/icustays.csv.gz"))
icustays_tble <- icustays_master  %>%
  print(width = Inf)
```

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays?  
**Solution:** I tabulated `icustays_tblethe` by `subject_id` and then obtained 53,150 unique `subject_id`'s, which is less than the total number of rows. Hence, `subject_id` can have multiple ICU stays.
```{r}
nrow(distinct(icustays_tble, subject_id))
```

3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.  
**Solution:** I used `arrange` to sort `intimes` in ascending order, and then I removed the duplicates for `subject_id`. As a result, the new table has 53,150 rows.  
```{r}
icustays_tble <- icustays_tble %>%
  arrange(subject_id, intime)  %>%
  distinct(subject_id, .keep_all = TRUE) %>%
  print(width = Inf)
```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.  
**Solution:** I used `read_csv` to import this file as a tibble.
```{r}
admissions_master <- read_csv(str_c(mimic_path,"/core/admissions.csv.gz"))
admissions_tble <- admissions_master %>%
  print(width = Inf)
```

2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadm_id`.  
**Solution:** I used `semi_join` to return the rows of `admissions_tble` with the same `subject_id` and `hadm_id` as `icustays_tble`.
```{r}
admissions_tble <- admissions_tble %>%
  arrange(subject_id, hadm_id) %>%
  semi_join(icustays_tble, by = c("subject_id", "hadm_id")) %>%
  print(width = Inf)
```

3. Summarize the following variables by graphics.   
**Solution:** Here is the code and output:

- admission year
```{r}
ggplot(data = admissions_tble) + 
  geom_bar(mapping = aes(x = year(admittime))) +
  labs(title = "Bar plot of admission time by year") +
  labs(x = "Admission year")
```
- admission month
```{r}
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = lubridate::month(admittime, label = T))) +
  labs(title = "Bar plot of admission time by month") +
  labs(x = "Admission month")
```
- admission month day
```{r}
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = mday(admittime))) +
  labs(title = "Bar plot of admission time by month and day") +
  labs(x = "Admission month day")
```
- admission week day  
```{r}
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = lubridate::wday(admittime, label = T))) +
  labs(title = "Bar plot of admission time by week day") +
  labs(x = "Admission week day")
```
- admission hour (anything unusual?)
```{r}
ggplot(data = admissions_tble) + 
  stat_count(mapping = aes(x = hour(admittime))) +
  labs(title = "Bar plot of admission time by hour") +
  labs(x = "Admission hour")
```
**Solution:** From the above graph, 0 AM and 7 AM appear to be unusual because the frequency is higher than other times in spite of a late night and an early morning.


## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).  
**Solution:** I used `read_csv` to import this file as a tibble. After matching, the subset contains 53,150 rows.
```{r}
patients_master <- read_csv(str_c(mimic_path,"/core/patients.csv.gz"))
patients_tble <- patients_master
patients_tble <- patients_tble %>%
  semi_join(icustays_tble, by = c("subject_id")) %>%
  print(width = Inf)
```

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.  
**Solution:** I used a box plot and a histogram to visualize the distribution of the two variables.  
The boxplot indicates:
- Both distributions are skewed to the left and have almost the same range;
- The mean age of women is slightly higher than that of men;  
The histogram indicates:
- Many ages have more males than females;
- There is a significant number of patients with the maximum age (91 years).
```{r}
ggplot(data = patients_tble, mapping = aes(x = gender, y = anchor_age)) + 
  stat_boxplot(geom = "errorbar", width = 0.2) + 
  geom_boxplot()  +
  labs(title = "Distribution for patient’s age by gender (Boxplot)") +
  labs(y = "Patient’s age")

ggplot(data = patients_tble) + 
  geom_bar(mapping = aes(x = anchor_age, fill = gender)) +
  labs(title = "Distribution for patient’s age by gender (Histogram)") +
  labs(x = "Patient’s age")
max(patients_tble$anchor_age)
```

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
  ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.  
**Solution:** There are 122,103,667 rows in `labevents.csv.gz`. (30,198,826 rows in `labevents_filtered_itemid.csv.gz`.)
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | tail -n +2 | wc -l"
  ), 
  intern = TRUE
)
```

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`.  

Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.

**Solution:** I used `read_csv` to import this file as a tibble and obtain 30,198,826 rows. The subset after narrowing has 16,698,462 rows.
```{r}
labevents_master <- 
  read_csv(str_c(mimic_path,"/hosp/labevents_filtered_itemid.csv.gz"), 
           col_types = cols_only(subject_id = col_double(), 
                                 itemid = col_double(), 
                                 charttime = col_datetime(), 
                                 valuenum = col_double()),
           lazy = TRUE)

labevents_tble <- labevents_master

d_labitems_tble <- read_csv(str_c(mimic_path,"/hosp/d_labitems.csv.gz"))

# item list we plan to use (but we do not need this anymore)
choice_lab <- c("50912", "50971", "50983", "50902", "50882", 
                "51221", "51301", "50931", "50960", "50893")

labevents_tble <- labevents_tble %>%
  # only contain these items for the patients in icustays_tble
  semi_join(icustays_tble, by = c("subject_id")) %>%
  # add label in d_labitems
  left_join(select(d_labitems_tble, itemid, label), by = c("itemid")) %>%
  print(width = Inf)
```

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 
**Solution:** The following code made `labevents_tble` 523252 rows.
```{r}
if (file.exists("labevents_tble.rds")) {
  labevents_tble <- read_rds("labevents_tble.rds")
} else {
  labevents_tble <- labevents_tble %>%
    # add intime from icustays_tble to find the first lab measurement
    left_join(select(icustays_tble, subject_id, intime), 
              by = c("subject_id")) %>%
    # only keep lab measurements after intime
    filter(charttime >= intime) %>%
    # sort charttime in ascending order by subject id and item id
    group_by(subject_id, itemid) %>%
    arrange(charttime, .by_group = TRUE) %>%
    # only keep the first lab measurement by group
    slice_head(n = 1) %>%
    # spread label
    ungroup() %>%
    select(-c(itemid, charttime, intime)) %>%
    pivot_wider(names_from = label, values_from = valuenum) %>%
    print(width = Inf) %>%
    write_rds("labevents_tble.rds")
}
```

4. Summarize the lab measurements by appropriate numerics and graphics.  
**Solution:**
```{r}
# numerical summary
summary(labevents_tble[-1])

# graphical summary
labevents_tble %>%
  select(2:11) %>%
  gather() %>%
  group_by(key) %>%
  filter(value > quantile(value, 0.025, na.rm = TRUE) 
         & value < quantile(value, 0.975, na.rm = TRUE)) %>%
  ungroup %>%
  ggplot() +
  geom_boxplot(mapping = aes(y = value)) +
  facet_wrap(~key, scales = "free_y") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) 
```

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.  
**Solution:** I used `read_csv` to import this file as a tibble and obtain 23,679,058 rows. The subset kept all the rows.
```{r}
chartevents_master <-
  read_csv(str_c(mimic_path,"/icu/chartevents_filtered_itemid.csv.gz"),
           col_types = cols_only(subject_id = col_double(),
                                 hadm_id = col_double(),
                                 itemid = col_double(),
                                 itemid = col_double(), 
                                 charttime = col_datetime(), 
                                 valuenum = col_double()),
           lazy = TRUE)

chartevents_tble <- chartevents_master

d_items_tble <- read_csv(str_c(mimic_path,"/icu/d_items.csv.gz"))

# item list we plan to use (but we do not need this anymore)
choice2 <- c("220045", "220181", "220179", "223761", "220210")

chartevents_tble <- chartevents_tble %>%
  # only containing these items for the patients in icustays_tble
  semi_join(icustays_tble, by = c("subject_id")) %>%
  # add label in d_items
  left_join(select(d_items_tble, itemid, label), by = c("itemid")) %>%
  print(width = Inf)
```

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 
```{r}
if (file.exists("chartevents_tble.rds")) {
  chartevents_tble <- read_rds("chartevents_tble.rds")
} else {
  chartevents_tble <- chartevents_tble %>%
    # add intime from icustays_tble to find the first lab measurement
    left_join(select(icustays_tble, subject_id, intime), 
              by = c("subject_id")) %>%
    # only keep lab measurements after intime
    filter(charttime >= intime) %>%
    # sort charttime in ascending order by subject id and item id
    group_by(subject_id, itemid) %>%
    arrange(charttime, .by_group = TRUE) %>%
    # only keep the first lab measurement by group
    slice_head(n = 1) %>%
    # spread label
    ungroup() %>%
    select(c(subject_id, label, valuenum)) %>%
    pivot_wider(names_from = label, values_from = valuenum) %>%
    print(width = Inf) %>%
    write_rds("chartevents_tble.rds")
}
```

3. Summarize these vital measurements by appropriate numerics and graphics. 
```{r}
# numerical summary
summary(chartevents_tble[-1])

# graphical summary
chartevents_tble %>%
  select(2:6) %>%
  gather() %>%
  group_by(key) %>%
  filter(value > quantile(value, 0.025, na.rm = TRUE) 
         & value < quantile(value, 0.975, na.rm = TRUE)) %>%
  ungroup %>%
  ggplot() +
  geom_boxplot(mapping = aes(y = value)) +
  facet_wrap(~key, scales = "free_y") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)
```{r}
if (file.exists("mimic_icu_cohort.rds")) {
  mimic_icu_cohort <- read_rds("mimic_icu_cohort.rds")
} else {
mimic_icu_cohort <- icustays_tble %>%
  left_join(patients_tble, by = c("subject_id")) %>%
  left_join(admissions_tble, by = c("subject_id", "hadm_id")) %>%
  # compute age at admission
  mutate(age_hadm = anchor_age + year(admittime) - anchor_year) %>%
  # keep only patients who are over 18
  filter(age_hadm > 18) %>%
  left_join(labevents_tble, by = c("subject_id")) %>%
  left_join(chartevents_tble, by = c("subject_id")) %>%
  # create thirty_day_mort that has 1 if a patient died <= 30 days and 0 if not
  mutate(thirty_day_mort = 
           ifelse(is.na(deathtime), Inf, 
                  ifelse(as.Date(deathtime) - as.Date(admittime) <= 30, "Yes",
                         "No"))) %>%
  print(width = Inf) %>%
  write_rds("mimic_icu_cohort.rds")
}
```

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)
```{r}
summary(mimic_icu_cohort[-1])
table(mimic_icu_cohort$thirty_day_mort)

col_names <- c("ethnicity", "language", "insurance", "marital_status", "gender")

for (i in 1:3) {
  mimic_icu_cohort %>%
    select(ethnicity, language, thirty_day_mort) %>%
    ggplot() +
    geom_bar(mapping = aes(x = thirty_day_mort, fill = mimic_icu_cohort[1]), position = "fill") +
    labs(y = "percent") +
    scale_x_discrete(limit = c("No", "Yes")) +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Title Xxxx") +
    labs(x = "30 day mortality") +
    coord_flip() 
}

mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = ethnicity), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Title Xxxx") +
  labs(x = "30 day mortality") +
  coord_flip() 

mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = language), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Title Xxxx") +
  labs(x = "30 day mortality") +
  coord_flip() 

mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = insurance), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Title Xxxx") +
  labs(x = "30 day mortality") +
  coord_flip() 

mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = marital_status), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Title Xxxx") +
  labs(x = "30 day mortality") +
  coord_flip()

mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = gender), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Title Xxxx") +
  labs(x = "30 day mortality") +
  coord_flip()


ggplot(data = mimic_icu_cohort) + 
  geom_boxplot(mapping = aes(x = thirty_day_mort, y = age_hadm)) +
  scale_x_discrete(limit = c("No", "Yes")) +
  coord_flip() 

mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = ethnicity), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Title Xxxx") +
  labs(x = "30 day mortality") +
  coord_flip() 

mimic_icu_cohort %>%
  select(ethnicity, language, insurance, marital_status, gender, thirty_day_mort) %>%
  gather(-thirty_day_mort, key="key", value="value") %>% 
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = value), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Title Xxxx") +
  labs(x = "30 day mortality") +
  facet_grid(~key, scales = "free") +
  coord_flip() 
```

- `thirty_day_mort` vs first lab measurements
```{r}
# graphical summary
mimic_icu_cohort %>%
  gather(28:37, key="key", value="value") %>%
  group_by(key) %>%
  filter(value > quantile(value, 0.025, na.rm = TRUE) 
         & value < quantile(value, 0.975, na.rm = TRUE)) %>%
  ungroup %>%
  ggplot(mapping = aes(x = thirty_day_mort, y = value)) +
  geom_boxplot() +
  scale_x_discrete(limit = c("Yes", "No")) +
  labs(x = "30 day mortality") +
  facet_wrap(~key, scales = "free_y") +
  labs(title = "Distribution for first lab measurements by 30 day mortality")
```

- `thirty_day_mort` vs first vital measurements
```{r}
# graphical summary
mimic_icu_cohort %>%
  gather(38:42, key="key", value="value") %>%
  group_by(key) %>%
  filter(value > quantile(value, 0.025, na.rm = TRUE) 
         & value < quantile(value, 0.975, na.rm = TRUE)) %>%
  ungroup %>%
  ggplot(mapping = aes(x = thirty_day_mort, y = value)) +
  geom_boxplot() +
  scale_x_discrete(limit = c("Yes", "No")) +
  labs(x = "30 day mortality") +
  facet_wrap(~key, scales = "free_y") +
  labs(title = "Distribution for first vital measurements by 30 day mortality")
```

- `thirty_day_mort` vs first ICU unit
```{r}
# graphical summary
mimic_icu_cohort
mimic_icu_cohort %>%
  ggplot() +
  geom_bar(mapping = aes(x = thirty_day_mort, fill = first_careunit), position = "fill") +
  labs(y = "percent") +
  scale_x_discrete(limit = c("Yes", "No")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion for the first ICU unit by 30 day mortality") +
  labs(x = "30 day mortality", fill ="First ICU unit")
```
